---
title: "Bike Sharing Case Study"
knit: (function(input_file, encoding) {
    out_dir <- 'docs';
    rmarkdown::render(input_file,
      encoding=encoding,
      output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Robert Martin Sim"
output: html_document
---

### Scenario

A junior data analyst working is part of the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director of marketing believes the company's future success depends on maximizing the number of annual memberships. Therefore, the team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, the team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve the analyst's recommendations, so they must be backed up with compelling data insights and professional data visualizations.

### Analysis Process

### 1. Loading Required Packages and Data

The packages contains the functions that will be used in this study. If they cannot be loaded, the packages must be installed first.

```{r Load packages}
library(dplyr)
library(tidyverse)
library(janitor)
library(skimr)
library(lubridate)
```

##### 1.1 Import dataframes of 2022 data

The project officially begins with importing all the needed data.

```{r Datasets of 2022}
jan_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202201-divvy-tripdata.csv")
feb_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202202-divvy-tripdata.csv")
mar_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202203-divvy-tripdata.csv")
apr_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202204-divvy-tripdata.csv")
may_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202205-divvy-tripdata.csv")
jun_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202206-divvy-tripdata.csv")
jul_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202207-divvy-tripdata.csv")
aug_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202208-divvy-tripdata.csv")
sep_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202209-divvy-tripdata.csv")
oct_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202210-divvy-tripdata.csv")
nov_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202211-divvy-tripdata.csv")
dec_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202212-divvy-tripdata.csv")
```

### 2. Wrangle Data and Combine Into a Single Datset

##### 2.2 Compare column of dataframes

The column names of each uploaded datasets are then compared.

```{r Call out column names}
colnames(jan_2022)
colnames(feb_2022)
colnames(mar_2022)
colnames(apr_2022)
colnames(may_2022)
colnames(jun_2022)
colnames(jul_2022)
colnames(aug_2022)
colnames(sep_2022)
colnames(oct_2022)
colnames(nov_2022)
colnames(dec_2022)
```

In this step, it could be seen that they possess the same column names.

##### 2.3 Compare datatype of each dataframe using loop

```{r Call out datatypes of each datasets}
months_2022 <- list(jan_2022, feb_2022, mar_2022, apr_2022, may_2022, jun_2022, jul_2022, aug_2022, sep_2022, oct_2022, nov_2022, dec_2022)
results <- lapply(months_2022, function(month) Map(function(a, b) all(class(a) == class(b)), jan_2022, month))
print(results)
```

In this step, it could be seen that the dataframes have the same datatype for each of their columns.

##### 2.4 Stack individual dataframes into one dataframe

```{r Bind all into one dataset}
all_trips <- bind_rows(jan_2022, feb_2022, mar_2022, apr_2022, may_2022, jun_2022, jul_2022, aug_2022, sep_2022, oct_2022, nov_2022, dec_2022)
```

The individual dataframes are compiled into one big dataframe to start doing the needed cleaning and calculations.

### 3. Clean up and Add Data to Prepare for Analysis

```{r Check state of combined dataset}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
tail(all_trips)
str(all_trips)
summary(all_trips)
```

The new table that has been created is inspected.

##### 3.1 Add year, month and day columns

```{r Extract year, month, day from started_at column}
all_trips$year <- format(as.Date(all_trips$started_at), "%Y")
all_trips$month <- format(as.Date(all_trips$started_at), "%m")
all_trips$day <- format(as.Date(all_trips$started_at), "%d")
```

Information regarding the year, month and day of the trips are extracted to three new columns.

##### 3.2 Add day_of_week column in making day as number ( 1 = Sunday)

```{r Convert days to numeric}
all_trips <- all_trips %>% mutate(day_of_the_week = wday(as.Date(all_trips$started_at)))
```

The *day_of_week* column is transformed to numbers with Sunday being "1".

##### 3.3 Add month_of_year column in making month as number ( 1 = January)

```{r Convert months to numeric}
all_trips <- all_trips %>% mutate(month_of_year = month(as.Date(all_trips$started_at)))
```

The *month_of_year* column is transformed to numbers with January being "1".

##### 3.4 Adding column ride_length

```{r Calculate duration and set to *ride_length*}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

The *ride_length* column is added to calculate the duration of each trips taken.

##### 3.5 Convert column ride_length to numeric

```{r *ride_length* to numeric}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

The *ride_length* column is converted to numeric type.

##### 3.4 Remove negative ride_length

```{r Remove negative from column}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

Negative values are removed from the *ride_length* column since it is not possible to have trip durations lesser than 0.

### 4. Conduct Descriptive Analysis

##### 4.1 ride_length descriptive analysis (in sec)

```{r Analysis of column}
summary(all_trips_v2$ride_length)
```

The minimum, maximum, median, mean, 1st quartile and 3rd quartile and the number of NA values of the *ride_length* column is calculated.

##### 4.2 Compare members and casual users

```{r Compare column}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

The mean, median, maximum and minimum values for members (annual) and casual members are compared with each other.

##### 4.3 Compare average ride_length by each day for each members vs casual

```{r Compare averages}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_the_week, FUN = mean)
```

The average trip duration in each day of the day of the week for each annual and casual members are compared with each other.

##### 4.4 Visualize number by rides by rider type AND drop NA values; days_of_the_week VS number of rides

```{r Plot bar chart for days_of_the_week VS number of rides}
all_trips_v2 %>%
  drop_na() %>%
  group_by(member_casual,day_of_the_week) %>% 
  summarise(number_of_rides=n(), average_duration = mean(ride_length)) %>% 
  group_by(day_of_the_week) %>% 
  mutate(percent_number_of_rides = number_of_rides/sum(number_of_rides)) %>% 
  mutate(percent_number_of_rides = round(percent_number_of_rides*100)) %>%
  ggplot(aes(x = day_of_the_week, y = number_of_rides, fill= member_casual)) + 
  geom_col(position="stack") + # Add stacked column graph
  scale_x_continuous(breaks = seq(1, 7)) + # Set x-axis to 1-7
  geom_text(aes(label = paste0(percent_number_of_rides, "%")), 
            position = position_stack(vjust = 0.5)) + # Add percentage labels for subscription types
  ggtitle("Subscription Type on Number of rides per Day of Week", subtitle = "(1 = Sunday)") +
  xlab("Day of the Week") +
  ylab("Number of Rides") +
  scale_fill_discrete(name = "Subscription Type") + # Change legend title for member_casual
  theme(title = element_text(size = 15),
        axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_blank())
```

The number of rides per customer type is plotted against the day of each week and is visualized using a stacked column chart. The percentages are shown for each member (subscription) type. The findings are as follows:

-   Saturday has the most number of rides.

-   Monday has the least number of rides.

-   Members have more number of rides per day except in Saturday.

-   Members have the most number of rides in Tuesday and Wednesday and least in Saturday.

-   Casuals have the most number of rides in Saturday and least in Tuesday and Wednesday.

##### 4.5 Visualize average duration; days_of_the_week VS ride_length

```{r Plot bar chart for days_of_the_week VS ride_length}
all_trips_v2 %>%
  drop_na() %>%
  group_by(member_casual,day_of_the_week) %>% 
  summarise(number_of_rides=n(), average_duration = mean(ride_length)) %>% 
  group_by(day_of_the_week) %>% 
  mutate(percent_average_duration = average_duration/sum(average_duration)) %>% 
  mutate(percent_average_duration = round(percent_average_duration*100)) %>%
  ggplot(aes(x = day_of_the_week, y = average_duration, fill=member_casual)) + 
  geom_col(position="stack") +
  scale_x_continuous(breaks = seq(1, 7)) +
  geom_text(aes(label = paste0(percent_average_duration, "%")), 
            position = position_stack(vjust = 0.5)) +
  ggtitle("Subscription Type on Ride Duration per Day of Week", subtitle = "(1 = Sunday)") +
  xlab("Day of Week") +
  ylab("Ride Duration") +
  scale_fill_discrete(name = "Subscription Type") + 
  theme(title = element_text(size = 15),
          axis.title.x = element_text(size = 15),
          axis.title.y = element_text(size = 15),
          axis.text.x = element_text(size = 12),
          axis.text.y = element_blank())
```

The ride duration per customer type is plotted against the day of each week and is visualized using a stacked column chart. The percentages are shown for each member (subscription) type. The findings are as follows:

-   Sunday and Saturday both have the highest ride duration.

-   Wednesday has the lowest ride duration.

-   Casuals have higher ride duration than Members in every day of the week.

##### 4.6 Visualize number of rides by subscription type; month_of_year VS number of rides

```{r Plot bar chart for month_of_year VS number of rides}
all_trips_v2 %>%
  drop_na() %>%
  group_by(member_casual, month_of_year) %>% 
  summarise(number_of_rides=n(), average_duration = mean(ride_length)) %>% 
  group_by(month_of_year) %>% 
  mutate(percent_number_of_rides = number_of_rides/sum(number_of_rides)) %>% 
  mutate(percent_number_of_rides = round(percent_number_of_rides*100)) %>% 
  ggplot(aes(x = month_of_year, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "stack") +
  scale_x_continuous(breaks = seq(1, 12)) +
  geom_text(aes(label = paste0(percent_number_of_rides, "%")), 
            position = position_stack(vjust = 0.5)) +
  ggtitle("Subscription Type on Number of rides per Month of Year", subtitle = "(1 = January)") +
  xlab("Month of Year") +
  ylab("Number of Rides") +
  scale_fill_discrete(name = "Subscription Type") +
  theme(title = element_text(size = 15),
        axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_blank())
```

The number of rides per customer type is plotted against the month of the year and is visualized using a stacked column chart. The percentages are shown for each member (subscription) type. The findings are as follows:

-   July has the most number of rides.

-   January has the least number of rides.

-   Members and Casuals have the most number of rides in June, July and August and the least in January and February.

##### 4.7 Visualize average duration; month_of_year VS ride_length

```{r Plot bar chart for month_of_year VS ride_length}
all_trips_v2 %>%
  drop_na() %>%
  group_by(member_casual,month_of_year) %>% 
  summarise(number_of_rides=n(), average_duration = mean(ride_length)) %>% 
  group_by(month_of_year) %>% 
  mutate(percent_average_duration = average_duration/sum(average_duration)) %>% 
  mutate(percent_average_duration = round(percent_average_duration*100)) %>%
  ggplot(aes(x = month_of_year, y = average_duration, fill=member_casual)) +
  geom_col(position="stack") +
  scale_x_continuous(breaks = seq(1, 12)) +
  geom_text(aes(label = paste0(percent_average_duration, "%")), 
            position = position_stack(vjust = 0.5)) +
  ggtitle("Subscription Type on Ride Duration per Month of Year", subtitle = "(1 = January)") +
  xlab("Month of Year") +
  ylab("Ride Duration") +
  scale_fill_discrete(name = "Subscription Type") +
  theme(title = element_text(size = 15),
        axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_blank())
```

The ride duration per customer type is plotted against the month of the year and is visualized using a stacked column chart. The percentages are shown for each member (subscription) type. The findings are as follows:

-   Casuals have higher ride duration than Members in every month of the year.

-   April and May have the highest ride duration while December has the lowest ride duration.

##### 4.8 Visualize number of rides; bike type vs number of rides

```{r Plot bar chart for bike_type VS number _of_rides}
all_trips_v2 %>%
  drop_na() %>%
  mutate(rideable_type = gsub("_", " ", rideable_type)) %>% # Replace all "_" to " " in rideable_type
  group_by(member_casual, rideable_type) %>% 
  summarise(number_of_rides=n(), average_duration = mean(ride_length)) %>% 
  group_by(rideable_type) %>% 
  mutate(percent_number_of_rides = number_of_rides/sum(number_of_rides)) %>% 
  mutate(percent_number_of_rides = round(percent_number_of_rides*100)) %>% 
  ggplot(aes(x = rideable_type, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "stack") +
  geom_text(aes(label = paste0(percent_number_of_rides, "%")), 
            position = position_stack(vjust = 0.5)) +
  ggtitle("Subscription Type on Number of rides per Bike Type") +
  xlab("Bike Type") +
  ylab("Number of Rides") +
  scale_fill_discrete(name = "Subscription Type") +
  theme(title = element_text(size = 15),
      axis.title.x = element_text(size = 15),
      axis.title.y = element_text(size = 15),
      axis.text.x = element_text(size = 12),
      axis.text.y = element_blank())
```

The number of rides per customer type is plotted against the bike type and is visualized using a stacked column chart. The percentages are shown for each member (subscription) type. The findings are as follows:

-   Classic bike has highest number of rides. Members rode classic bikes more than Casuals.

-   Docked bikes has lowest number of rides. Docked bikes are ridden only by casuals.

-   Members rode electric bikes more than Casuals.

##### 4.9 Visualize average duration; bike type vs ride_length

```{r Plot bar chart for bike_type VS ride_length}
all_trips_v2 %>%
  drop_na() %>%
  mutate(rideable_type = gsub("_", " ", rideable_type)) %>%
  group_by(member_casual, rideable_type) %>% 
  summarise(number_of_rides=n(), average_duration = mean(ride_length)) %>% 
  group_by(rideable_type) %>% 
  mutate(percent_average_duration = average_duration/sum(average_duration)) %>% 
  mutate(percent_average_duration = round(percent_average_duration*100)) %>%
  ggplot(aes(x = rideable_type, y = average_duration, fill = member_casual)) +
  geom_col(position = "stack") +
  geom_text(aes(label = paste0(percent_average_duration, "%")), 
            position = position_stack(vjust = 0.5)) +
  ggtitle("Subscription Type on Ride Duration per Bike Type") +
  xlab("Bike Type") +
  ylab("Ride Duration") +
  scale_fill_discrete(name = "Subscription Type") +
  theme(title = element_text(size = 15),
        axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_blank());
```

The ride duration per customer type is plotted against the bike type and is visualized using a stacked column chart. The percentages are shown for each member (subscription) type. The findings are as follows:

-   Docked bikes have the highest ride duration which are only ridden by Casuals.

-   Electric bikes have the lowest ride duration.

-   Casuals have higher ride duration than Members per bike type.

### 5. Export Summary File into csv

```{r Summary of datasets exported to csv}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual +
                      all_trips_v2$day_of_the_week +
                      all_trips_v2$month_of_year+
                      all_trips_v2$rideable_type, FUN = mean)
write.csv(counts, file='~/Bike-Sharing/avg_ride_length.csv')
```

The dataframe used in analysis is then exported to a csv file in the event that it will be used for analysis in other platforms.

### Summary and Recommendations

From the analysis done, the following actions are recommended to convert more casual members to annual type subscription:

-   Regarding the bike type, promotions should be offered for docked bikes as they are more popular among casual members in terms of both ride frequency and duration, compared to annual members.

-   Regarding the monthly promotions, July should be favored as it has the highest number of rides. Another promotion should be considered based on trip duration, as casual members consistently have longer rides compared to annual members every month.

-   Regarding the weekly promotions, Saturdays should receive a promotion aimed at casual members as this is the day when casual members have the highest number of rides and longest trip duration.
