---
title: "Bike Sharing Case Study"
knit: (function(input_file, encoding) {
    out_dir <- 'docs';
    rmarkdown::render(input_file,
      encoding=encoding,
      output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Robert Martin Sim"
output: html_document
---

<br>

### Loading Required Packages

The packages contains the functions that will be used in this project. If they cannot be loaded, the packages must be installed first.

```{r Load packages}
library(dplyr)
library(tidyverse)
library(janitor)
library(skimr)
library(lubridate)
```

### 1. Collect Data 

The project officially begins with importing all the needed data. 

#### Import datasets of 2022 data

```{r Datasets of 2022}
jan_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202201-divvy-tripdata.csv")
feb_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202202-divvy-tripdata.csv")
mar_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202203-divvy-tripdata.csv")
apr_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202204-divvy-tripdata.csv")
may_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202205-divvy-tripdata.csv")
jun_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202206-divvy-tripdata.csv")
jul_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202207-divvy-tripdata.csv")
aug_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202208-divvy-tripdata.csv")
sep_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202209-divvy-tripdata.csv")
oct_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202210-divvy-tripdata.csv")
nov_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202211-divvy-tripdata.csv")
dec_2022 <- read_csv("~/Bike-Sharing/Case Study - Bike-Share/Datasets/2022/202212-divvy-tripdata.csv")
```

### 2. Wrangle Data and Combine Into a Single Datset

##### 2.2 Compare column of datasets

Column names of each uploaded datasets are then compared.

```{r Call out column names}
colnames(jan_2022)
colnames(feb_2022)
colnames(mar_2022)
colnames(apr_2022)
colnames(may_2022)
colnames(jun_2022)
colnames(jul_2022)
colnames(aug_2022)
colnames(sep_2022)
colnames(oct_2022)
colnames(nov_2022)
colnames(dec_2022)
```

It could be seen that they possess the same column names.

##### 2.3 Compare datatype of each datasets using loop

```{r Call out datatypes of each datasets}
months_2022 <- list(jan_2022, feb_2022, mar_2022, apr_2022, may_2022, jun_2022, jul_2022, aug_2022, sep_2022, oct_2022, nov_2022, dec_2022)
results <- lapply(months_2022, function(month) Map(function(a, b) all(class(a) == class(b)), jan_2022, month))
print(results)
```

##### 2.4 Stack individual dateset into one dataset

```{r Bind all into one dataset}
all_trips <- bind_rows(jan_2022, feb_2022, mar_2022, apr_2022, may_2022, jun_2022, jul_2022, aug_2022, sep_2022, oct_2022, nov_2022, dec_2022)
```

### 3. Clean up and Add Data to Prepare for Analysis

```{r Check state of combined dataset}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
tail(all_trips)
str(all_trips)
summary(all_trips)
```

##### 3.1 Add year, month and day columns

```{r Extract year, month, day from started_at column}
all_trips$year <- format(as.Date(all_trips$started_at), "%Y")
all_trips$month <- format(as.Date(all_trips$started_at), "%m")
all_trips$day <- format(as.Date(all_trips$started_at), "%d")
```

##### 3.2 Add day_of_week column in making day as number ( 1 = Sunday)

```{r Convert days to numeric}
all_trips <- all_trips %>% mutate(day_of_the_week = wday(as.Date(all_trips$started_at)))
```

##### 3.3 Add month_of_year column in making month as number ( 1 = January)

```{r Convert months to numeric}
all_trips <- all_trips %>% mutate(month_of_year = month(as.Date(all_trips$started_at)))
```

##### 3.4 Adding column ride_length

```{r Calculate duration and set to *ride_length*}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

##### 3.5 Convert column ride_length to numeric

```{r *ride_length* to numeric}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

##### 3.4 Remove negative ride_length

```{r Remove negative from column}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

### 4. Conduct Descriptive Analysis

##### 4.1 ride_length descriptive analysis (in sec)

```{r Analysis of column}
summary(all_trips_v2$ride_length)
```

##### 4.2 Compare members and casual users

```{r Compare column}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

##### 4.3 Compare average ride_length by each day for each members vs casual

```{r Compare averages}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_the_week, FUN = mean)
```

##### 4.4 Visualize number by rides by rider type AND drop NA values; days_of_the_week VS number of rides

```{r Plot bar chart for days_of_the_week VS number of rides}
all_trips_v2 %>%
  drop_na() %>%
  group_by(member_casual,day_of_the_week) %>% 
  summarise(number_of_rides=n(), average_duration = mean(ride_length)) %>% 
  group_by(day_of_the_week) %>% 
  mutate(percent_number_of_rides = number_of_rides/sum(number_of_rides)) %>% 
  mutate(percent_number_of_rides = round(percent_number_of_rides*100)) %>%
  ggplot(aes(x = day_of_the_week, y = number_of_rides, fill= member_casual)) + 
  geom_col(position="stack") + # Add stacked column graph
  scale_x_continuous(breaks = seq(1, 7)) + # Set x-axis to 1-7
  geom_text(aes(label = paste0(percent_number_of_rides, "%")), 
            position = position_stack(vjust = 0.5)) + # Add percentage labels for subscription types
  ggtitle("Subscription Type on Number of rides per Day of Week", subtitle = "(1 = Sunday)") +
  xlab("Day of the Week") +
  ylab("Number of Rides") +
  scale_fill_discrete(name = "Subscription Type") + # Change legend title for member_casual
  theme(title = element_text(size = 15),
        axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_blank())
```

##### 4.5 Visualize average duration; days_of_the_week VS ride_length

```{r Plot bar chart for days_of_the_week VS ride_length}
all_trips_v2 %>%
  drop_na() %>%
  group_by(member_casual,day_of_the_week) %>% 
  summarise(number_of_rides=n(), average_duration = mean(ride_length)) %>% 
  group_by(day_of_the_week) %>% 
  mutate(percent_average_duration = average_duration/sum(average_duration)) %>% 
  mutate(percent_average_duration = round(percent_average_duration*100)) %>%
  ggplot(aes(x = day_of_the_week, y = average_duration, fill=member_casual)) + 
  geom_col(position="stack") +
  scale_x_continuous(breaks = seq(1, 7)) +
  geom_text(aes(label = paste0(percent_average_duration, "%")), 
            position = position_stack(vjust = 0.5)) +
  ggtitle("Subscription Type on Ride Duration per Day of Week", subtitle = "(1 = Sunday)") +
  xlab("Day of Week") +
  ylab("Ride Duration") +
  scale_fill_discrete(name = "Subscription Type") + 
  theme(title = element_text(size = 15),
          axis.title.x = element_text(size = 15),
          axis.title.y = element_text(size = 15),
          axis.text.x = element_text(size = 12),
          axis.text.y = element_blank())
```

##### 4.6 Visualize number of rides by subscription type; month_of_year VS number of rides

```{r Plot bar chart for month_of_year VS number of rides}
all_trips_v2 %>%
  drop_na() %>%
  group_by(member_casual, month_of_year) %>% 
  summarise(number_of_rides=n(), average_duration = mean(ride_length)) %>% 
  group_by(month_of_year) %>% 
  mutate(percent_number_of_rides = number_of_rides/sum(number_of_rides)) %>% 
  mutate(percent_number_of_rides = round(percent_number_of_rides*100)) %>% 
  ggplot(aes(x = month_of_year, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "stack") +
  scale_x_continuous(breaks = seq(1, 12)) +
  geom_text(aes(label = paste0(percent_number_of_rides, "%")), 
            position = position_stack(vjust = 0.5)) +
  ggtitle("Subscription Type on Number of rides per Month of Year", subtitle = "(1 = January)") +
  xlab("Month of Year") +
  ylab("Number of Rides") +
  scale_fill_discrete(name = "Subscription Type") +
  theme(title = element_text(size = 15),
        axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_blank())
```

##### 4.7 Visualize average duration; month_of_year VS ride_length

```{r Plot bar chart for month_of_year VS ride_length}
all_trips_v2 %>%
  drop_na() %>%
  group_by(member_casual,month_of_year) %>% 
  summarise(number_of_rides=n(), average_duration = mean(ride_length)) %>% 
  group_by(month_of_year) %>% 
  mutate(percent_average_duration = average_duration/sum(average_duration)) %>% 
  mutate(percent_average_duration = round(percent_average_duration*100)) %>%
  ggplot(aes(x = month_of_year, y = average_duration, fill=member_casual)) +
  geom_col(position="stack") +
  scale_x_continuous(breaks = seq(1, 12)) +
  geom_text(aes(label = paste0(percent_average_duration, "%")), 
            position = position_stack(vjust = 0.5)) +
  ggtitle("Subscription Type on Ride Duration per Month of Year", subtitle = "(1 = January)") +
  xlab("Month of Year") +
  ylab("Ride Duration") +
  scale_fill_discrete(name = "Subscription Type") +
  theme(title = element_text(size = 15),
        axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_blank())
```

##### 4.8 Visualize number of rides; bike type vs number of rides

```{r Plot bar chart for bike_type VS number _of_rides}
all_trips_v2 %>%
  drop_na() %>%
  mutate(rideable_type = gsub("_", " ", rideable_type)) %>% # Replace all "_" to " " in rideable_type
  group_by(member_casual, rideable_type) %>% 
  summarise(number_of_rides=n(), average_duration = mean(ride_length)) %>% 
  group_by(rideable_type) %>% 
  mutate(percent_number_of_rides = number_of_rides/sum(number_of_rides)) %>% 
  mutate(percent_number_of_rides = round(percent_number_of_rides*100)) %>% 
  ggplot(aes(x = rideable_type, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "stack") +
  geom_text(aes(label = paste0(percent_number_of_rides, "%")), 
            position = position_stack(vjust = 0.5)) +
  ggtitle("Subscription Type on Number of rides per Bike Type") +
  xlab("Bike Type") +
  ylab("Number of Rides") +
  scale_fill_discrete(name = "Subscription Type") +
  theme(title = element_text(size = 15),
      axis.title.x = element_text(size = 15),
      axis.title.y = element_text(size = 15),
      axis.text.x = element_text(size = 12),
      axis.text.y = element_blank())
```

##### 4.9 Visualize average duration; bike type vs ride_length

```{r Plot bar chart for bike_type VS ride_length}
all_trips_v2 %>%
  drop_na() %>%
  mutate(rideable_type = gsub("_", " ", rideable_type)) %>%
  group_by(member_casual, rideable_type) %>% 
  summarise(number_of_rides=n(), average_duration = mean(ride_length)) %>% 
  group_by(rideable_type) %>% 
  mutate(percent_average_duration = average_duration/sum(average_duration)) %>% 
  mutate(percent_average_duration = round(percent_average_duration*100)) %>%
  ggplot(aes(x = rideable_type, y = average_duration, fill = member_casual)) +
  geom_col(position = "stack") +
  geom_text(aes(label = paste0(percent_average_duration, "%")), 
            position = position_stack(vjust = 0.5)) +
  ggtitle("Subscription Type on Ride Duration per Bike Type") +
  xlab("Bike Type") +
  ylab("Ride Duration") +
  scale_fill_discrete(name = "Subscription Type") +
  theme(title = element_text(size = 15),
        axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_blank())
```

### 5. Export Summary File into csv

```{r Summary of datasets exported to csv}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual +
                      all_trips_v2$day_of_the_week +
                      all_trips_v2$month_of_year+
                      all_trips_v2$rideable_type, FUN = mean)
write.csv(counts, file='~/Bike-Sharing/avg_ride_length.csv')
```
